---
title: "Assignment 03 — In-Class: Calculating P-values Using a Permutation Test"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

> **In-class activity.** Complete this notebook during class and submit it before you leave.
> Add your comments and answers in the spaces marked with *[ Write your answer here. ]*

---

# Part 1 — Calculating a P-value Using a Permutation Test

> **Scenario:** A small clinical trial enrolled 40 patients. Twenty received a placebo
> (group 1, control) and twenty received a new drug (group 2, treatment). Each value
> represents that patient's change in symptom score. Higher scores mean worse symptoms.

---

## Step 0 — How the data were generated (reference only)

```{r step-0-data-generation, eval=FALSE}
### Reference only — do not run ###

# effect_size <- 0.25
# stddev      <- 0.5
# sample_size <- 20
#
# group1 <- round(rnorm(sample_size, mean = 0,           sd = stddev) * 10)
# group2 <- round(rnorm(sample_size, mean = effect_size, sd = stddev) * 10)
#
# print(group1)
# print(group2)
```

---

## Step 1 — Enter and inspect the data

```{r step-1-data}
# Define the two groups

group1 <- c( 0,  5,  2, -2,  8,
            -6,  0,  0, -6, -3,
            -7,  4, -2, -2,  0,
            -4, -1, -8,  0,  6)

group2 <- c(-2,  2, -2, -5,  7,
            11,  6,  2,  1,  1,
             6, -2, -1,  7,  7,
            -2,  4,  3, -1, 17)

sample_size <- length(group1)

cat("Group 1 (control)  — mean:", round(mean(group1), 2),
    " | sd:", round(sd(group1), 2), "\n")
cat("Group 2 (treatment)— mean:", round(mean(group2), 2),
    " | sd:", round(sd(group2), 2), "\n")
```

*[ Add your notes here. ]*

---

## Step 2 — Visualize the two distributions

```{r step-2-histograms}
# Overlapping histograms of the two groups
# rgb(r, g, b, alpha), add = TRUE, abline(v = ...)

hist(group1,
     col    = rgb(0, 0, 1, 0.25),
     breaks = seq(-20, 20, 2.5),
     xlim   = c(-20, 20),
     xlab   = "Change in symptom score",
     main   = "Symptom score distributions: control (blue) vs. treatment (red)")

hist(group2,
     col    = rgb(1, 0, 0, 0.25),
     breaks = seq(-20, 20, 2.5),
     add    = TRUE)

abline(v = mean(group1), col = "blue", lwd = 2)
abline(v = mean(group2), col = "red",  lwd = 2)

legend("topright",
       legend = c("Group 1 (control)", "Group 2 (treatment)"),
       fill   = c(rgb(0, 0, 1, 0.4), rgb(1, 0, 0, 0.4)),
       bty    = "n")
```

*[ Add your notes here. ]*

---

## Step 3 — Compute the observed test statistic

```{r step-3-test-statistic}
# Compute signal / noise (the test statistic)

signal <- mean(group2) - mean(group1)
noise  <- sqrt((var(group2) / sample_size) + (var(group1) / sample_size))

test_statistic <- signal / noise

cat("Signal (difference in means):", round(signal,         4), "\n")
cat("Noise  (SE of the difference):", round(noise,          4), "\n")
cat("Observed test statistic:      ", round(test_statistic, 4), "\n")
```

*[ Add your notes here. ]*

---

## Step 4 — Set up the permutation test

```{r step-4-setup}
# Set up containers for the permutation test

num_permutations         <- 10000
permuted_test_statistics <- numeric(num_permutations)
```

---

## Step 5 — Run the permutation loop

```{r step-5-permutation-loop}
# Permutation loop: shuffle the combined data, re-split, compute the test statistic

for (i in 1:num_permutations) {

  permuted_data <- sample(c(group1, group2))

  rand_group1 <- permuted_data[1:sample_size]
  rand_group2 <- permuted_data[(sample_size + 1):(2 * sample_size)]

  perm_signal <- mean(rand_group2) - mean(rand_group1)
  perm_noise  <- sqrt((var(rand_group2) / sample_size) + (var(rand_group1) / sample_size))

  permuted_test_statistics[i] <- perm_signal / perm_noise
}
```

*[ Add your notes here. ]*

---

## Step 6 — Visualize the null distribution

```{r step-6-null-distribution}
# Histogram of the null distribution; red dashed line = observed test statistic

hist(permuted_test_statistics,
     breaks = 60,
     col    = "lightsteelblue",
     border = "white",
     xlab   = "Test statistic under H0 (null distribution)",
     main   = "Null distribution from 10,000 permutations\n(dashed red = observed value)")

abline(v = test_statistic, col = "red", lwd = 2, lty = 2)
```

*[ Add your notes here. ]*

---

## Step 7 — Compute the permutation p-value

```{r step-7-pvalue}
# Compute the p-value as the fraction of permuted statistics >= observed

pvalue <- sum(permuted_test_statistics >= test_statistic) / num_permutations
cat("Permutation p-value:", pvalue, "\n")
```

*[ Add your notes here. ]*

---

## Step 8 — Compare to the parametric t-test

```{r step-8-ttest}
# Welch two-sample t-test for comparison

t.test(group2, group1, alternative = "greater")
```

---

## Questions

**Question 1:**
Using the language of this specific experiment (placebo vs. drug, symptom scores),
write a precise one-sentence definition of the permutation p-value you calculated.

*[ Write your answer here. ]*

---

**Question 2:**
Is the permutation p-value close to the t-test p-value? What does agreement
(or disagreement) between the two methods tell you?

*[ Write your answer here. ]*
